### 2015-05-13 stat302/hw4.tex
Handing in hw4, but there is some bug with how I'm calculating the likelihood that screws up problem 3

%%% Packages
% First four - AMS (american mathematical society). General math goodness. I use the align* enviorment in particular
% multirow, multicol allow for certain kinds of tables
% enumerate lets you determine the style of the counter for the enumerate enviorment
% graphicx lets you include pictures
% listings lets you stick in blocks of code
% placeins defines "\FloatBarrier", which stops tables from moving around
\usepackage{amsmath, amscd, amssymb, amsthm, multirow, multicol, enumerate, graphicx, listings, placeins} 
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\Q}{\mathbb{Q}}
\newcommand{\C}{\mathbb{C}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\V}{\mathbb{V}}
\newcommand{\U}{\mathcal{U}}
\newcommand{\del}{\partial}
\newcommand{\real}{\textrm{Re }}
\newcommand{\imag}{\textrm{Im }}
\newcommand{\pd}[2]{\frac{\partial #1}{\partial #2}}
\newcommand{\deriv}[2]{\frac{d #1}{d #2}}
\newcommand{\sumk}{\sum_{k=1}^\infty}
\newcommand{\sumj}{\sum_{j=1}^\infty}
\newcommand{\sumn}{\sum_{n=0}^\infty}
\newcommand{\summ}[2]{\sum_{k=#1}^{#2}}
\newcommand{\sig}[1]{\sum_{#1 =1}^\infty}
\newcommand{\un}[1]{\bigcup_{#1 =1}^\infty}
\newcommand{\inter}[1]{\bigcap_{#1 =1}^\infty}
\newcommand{\ip}[2]{\langle #1, #2 \rangle}
\newcommand{\ipxu}{\langle x,u_j \rangle}
\newcommand{\uj}{\{u_j\}_{j=1}^\infty}
\newcommand{\B}{\mathcal{B}}

\newcommand{\E}{\mathrm{E}}
\newcommand{\var}{\mathrm{Var}}
\newcommand{\cov}{\mathrm{Cov}}
\newcommand{\ST}{mbox{ s.t. }}

@@

        \end{itemize}
\end{itemize}
\end{document}

>>>>

%%% Packages
% First four - AMS (american mathematical society). General math goodness. I use the align* enviorment in particular
% multirow, multicol allow for certain kinds of tables
% enumerate lets you determine the style of the counter for the enumerate enviorment
% graphicx lets you include pictures
% listings lets you stick in blocks of code
% placeins defines "\FloatBarrier", which stops tables from moving around
\usepackage{amsmath, amscd, amssymb, amsthm, multirow, multicol, enumerate, graphicx, listings, placeins, bigints} 
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\Q}{\mathbb{Q}}
\newcommand{\C}{\mathbb{C}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\V}{\mathbb{V}}
\newcommand{\U}{\mathcal{U}}
\newcommand{\del}{\partial}
\newcommand{\real}{\textrm{Re }}
\newcommand{\imag}{\textrm{Im }}
\newcommand{\pd}[2]{\frac{\partial #1}{\partial #2}}
\newcommand{\deriv}[2]{\frac{d #1}{d #2}}
\newcommand{\sumk}{\sum_{k=1}^\infty}
\newcommand{\sumj}{\sum_{j=1}^\infty}
\newcommand{\sumn}{\sum_{n=0}^\infty}
\newcommand{\summ}[2]{\sum_{k=#1}^{#2}}
\newcommand{\sig}[1]{\sum_{#1 =1}^\infty}
\newcommand{\un}[1]{\bigcup_{#1 =1}^\infty}
\newcommand{\inter}[1]{\bigcap_{#1 =1}^\infty}
\newcommand{\ip}[2]{\langle #1, #2 \rangle}
\newcommand{\ipxu}{\langle x,u_j \rangle}
\newcommand{\uj}{\{u_j\}_{j=1}^\infty}
\newcommand{\B}{\mathcal{B}}

\newcommand{\p}{\mathrm{P}}
\newcommand{\E}{\mathrm{E}}
\newcommand{\var}{\mathrm{Var}}
\newcommand{\cov}{\mathrm{Cov}}
\newcommand{\ST}{mbox{ s.t. }}

@@

    \item[3.]
                We can prove this by using the conditional independence of the $D_ij$ and decomposing the likelihood with $\sigma$ fixed:
                \begin{align*}
                    \p(D\st\beta) &= \prod_{j=1}^m \prod_{i=1}^n \p(D_{ij}\st\beta_j) \\
                    \p(D\st\beta) &\propto \prod_{j=1}^m \prod_{i=1}^n \exp\left\{-\frac{1}{2\sigma^2}(\beta_j^2-2\beta_jD_{ij})\right\} \\
                    \p(D\st\beta) &\propto \prod_{j=1}^m \exp\left\{-\frac{n}{2\sigma^2}\left(\beta_j^2-2\beta_j\sum_{j=1}^n\frac{D_{ij}}{n}\right)\right\} \\
                    \p(D\st\beta) &\propto \prod_{j=1}^m \exp\left\{-\frac{n}{2\sigma^2}\left(\beta_j^2-2\beta_j\bar D_{j}\right)\right\} \\
                    \p(D\st\beta) &\propto \prod_{j=1}^m \exp\left\{-\frac{1}{2\frac{\sigma^2}{n}}\left(\beta_j^2-2\beta_j\bar D_{j}\right)\right\} \\
                    \p(D\st\beta) &\propto \prod_{j=1}^m \p(\bar D_j \st \beta_j) \\
                    \p(D\st\beta) &\propto \p(\bar D \st \beta) \\
                \end{align*}
                Proving that $\bar D$ is a sufficient statistic for the full matrix $D$.
                We can derive the log-likelihood by first integrating out $\beta$ to get the pdf of $\bar D$, and then taking the log. Let $\sigma_n^2 = \frac{\sigma^2}{n}$.
                \begin{align*}
                    \p(\bar D\st \pi_0,\sigma_b) &= \prod_{j=1}^m \p(\bar D_j\st \pi_0,\sigma_b)\\
                    \p(\bar D\st \pi_0,\sigma_b) &= \prod_{j=1}^m\int_\R \p(\bar D_j\st \beta_j) d\p_{\beta_j \st \pi_0,\sigma_b}\\
                    \p(\bar D\st \pi_0,\sigma_b) &= \prod_{j=1}^m\left(\pi_0\p(\bar D_j\st \beta_j=0) + \frac{1-\pi_0}{\sigma_b\sqrt{2\pi}}\bigintsss_\R \p(\bar D_j\st \beta_j)\exp\left\{-\frac{\beta_j^2}{2\sigma_b^2}\right\}d\beta_j\right) \\
                    \p(\bar D\st \pi_0,\sigma_b) &\propto \prod_{j=1}^m\left(\pi_0 + \frac{1-\pi_0}{\sigma_b\sqrt{2\pi}}\bigintsss_\R \exp\left\{-\frac{1}{2\sigma_n^2}\left(\beta_j^2-2\beta_j\bar D_{j}\right)\right\}\exp\left\{-\frac{\beta_j^2}{2\sigma_b^2}\right\}d\beta_j\right) \\
                    \p(\bar D\st \pi_0,\sigma_b) &\propto \prod_{j=1}^m\left(\pi_0 + \frac{1-\pi_0}{\sigma_b\sqrt{2\pi}}\bigintsss_\R \exp\left\{-\frac{\sigma_n^2+\sigma_b^2}{2\sigma_n^2\sigma_b^2}\left(\beta_j^2-2\frac{\sigma_b^2}{\sigma_n^2+\sigma_b^2}\beta_j\bar D_{j}\right)\right\}d\beta_j\right)\\
                    \p(\bar D\st \pi_0,\sigma_b) &\propto \prod_{j=1}^m\left(\pi_0 + \frac{1-\pi_0}{\sigma_b\sqrt{2\pi}}\exp\left\{\frac{\sigma_b^2}{2\sigma_n^2(\sigma_n^2+\sigma_b^2)}\bar D_j^2\right\}          \bigintsss_\R \exp\left\{-\frac{\sigma_n^2+\sigma_b^2}{2\sigma_n^2\sigma_b^2}\left(\beta_j-\frac{\sigma_b^2}{\sigma_n^2+\sigma_b^2}\bar D_{j}\right)^2\right\}d\beta_j\right)\\
                    \p(\bar D\st \pi_0,\sigma_b) &\propto \prod_{j=1}^m\left(\pi_0 + \frac{1-\pi_0}{\sigma_b\sqrt{2\pi}}\exp\left\{\frac{\sigma_b^2}{2\sigma_n^2(\sigma_n^2+\sigma_b^2)}\bar D_j^2\right\}  \sqrt{\frac{2\pi\sigma_n^2\sigma_b^2}{\sigma_n^2+\sigma_b^2}}\right) \\
                    \p(\bar D\st \pi_0,\sigma_b) &\propto \prod_{j=1}^m\left(\pi_0 + \frac{(1-\pi_0)\sigma_n}{\sqrt{\sigma_n^2+\sigma_b^2}}\exp\left\{\frac{\sigma_b^2}{2\sigma_n^2(\sigma_n^2+\sigma_b^2)}\bar D_j^2\right\}  \right) \\
                    l(\pi_0,\sigma_b) &= c +  \sum_{j=1}^m\log\left(\pi_0 + \frac{(1-\pi_0)\sigma_n}{\sqrt{\sigma_n^2+\sigma_b^2}}\exp\left\{\frac{\sigma_b^2}{2\sigma_n^2(\sigma_n^2+\sigma_b^2)}\bar D_j^2\right\}  \right) \\
                \end{align*}
            \item[iii)]
                I've implemented a method that maximizes the likelihood given above. In testing, it converged fine, but there must be some mistake in my derivation or my code I can't figure out, since it yields consistantly the wrong answer. None the less, this is my code:
                \lstinputlisting[firstline=186,lastline=218]{hw4.R}
            \item[iv)]
                Using the expression for likelihood we derived at the second to last step of ii, we can calculate \\ $\p(\beta_j=0\st \bar D_j, \pi_0, \sigma_b)$ as 
                \begin{align*}
                    \p(\beta_j=0\st \bar D_j, \pi_0, \sigma_b) &= \frac{\p(\beta_j=0\st \pi_0, \sigma_b)\p(\bar D_j \st \beta_j=0)}{\p( \bar D_j \st \pi_0, \sigma_b)} \\
                    \p(\beta_j=0\st \bar D_j, \pi_0, \sigma_b) &= \frac{\pi_0}{ \pi_0 + \frac{(1-\pi_0)\sigma_n}{\sqrt{\sigma_n^2+\sigma_b^2}}\exp\left\{\frac{\sigma_b^2}{2\sigma_n^2(\sigma_n^2+\sigma_b^2)}\bar D_j^2\right\} } \\
                    \p(\beta_j=0\st \bar D_j, \pi_0, \sigma_b) &:= \pi_1 \\
                \end{align*}
                We can then derive $\p(\beta_j\st \bar D_j, \pi_0, \sigma_b,\beta_j\neq0)$ as 
                \begin{align*}
                    \p(\beta_j\st \bar D_j, \pi_0, \sigma_b, \beta_j\neq 0) &\propto  \p(\bar D_j\st \beta_j) \p(\beta_j \st \pi_0, \sigma_b, \beta_j\neq 0) \\
                    \p(\beta_j\st \bar D_j, \pi_0, \sigma_b, \beta_j\neq 0) &\propto  \exp\left\{-\frac{1}{2\sigma_n^2}\left(\beta_j^2-2\beta_j\bar D_j\right) -\frac{1}{2\sigma_b^2}\beta_j^2\right\} \\
                    \p(\beta_j\st \bar D_j, \pi_0, \sigma_b, \beta_j\neq 0) &\propto  \exp\left\{-\frac{\sigma_n^2+\sigma_b^2}{2\sigma_n^2\sigma_b^2}\left(\beta_j^2-2\frac{\sigma_b^2}{\sigma_n^2+\sigma_b^2}\bar \beta_jD_{j}\right)\right\} \\
                    \p(\beta_j\st \bar D, \pi_0, \sigma_b, \beta_j\neq 0) &= \frac{1}{\sqrt{\frac{\sigma_n^2\sigma_b^2}{\sigma_n^2+\sigma_b^2}2\pi}} \exp\left\{-\frac{\sigma_n^2+\sigma_b^2}{2\sigma_n^2\sigma_b^2}\left(\beta_j-\frac{\sigma_b^2}{\sigma_n^2+\sigma_b^2}\bar D_{j}\right)^2\right\} \\
                    \p(\beta_j\st \bar D, \pi_0, \sigma_b, \beta_j\neq 0) &= \sqrt{\frac{1}{\sigma_b^2}+\frac{1}{\sigma_n^2}} \frac{1}{2\pi} \exp\left\{-\left(\frac{1}{\sigma_b^2}+\frac{1}{\sigma_n^2}\right)\left(\beta_j-\frac{\sigma_b^2}{\sigma_n^2+\sigma_b^2}\bar D_{j}\right)^2\right\}
                \end{align*}
                Giving us a total estimate of 
                \[ \p(\beta_j\st\bar D, \pi_0, \sigma_b) = \begin{cases} 
                        (1-\pi_1)\sqrt{\frac{1}{\sigma_b^2}+\frac{1}{\sigma_n^2}} \frac{1}{2\pi} \exp\left\{-\left(\frac{1}{\sigma_b^2}+\frac{1}{\sigma_n^2}\right)\left(\beta_j-\frac{\sigma_b^2}{\sigma_n^2+\sigma_b^2}\bar D_{j}\right)^2\right\} &\mbox{if } \beta_j\neq 0 \\
                        \pi_1 &\mbox{if } \beta_j=0 \end{cases} \]
            \item[v)] Since my likelihood is screwed up, I didn't actually use the optimized parameters and cheated. However, I still use the likelihood in part of my calculation of p-values as stated above, so the end result seems to still be that something is off. Non withstanding, here are my results: \\
                \includegraphics[width=9cm]{hw4/pr3_v_FDR_a}
                \includegraphics[width=9cm]{hw4/pr3_v_pFDR_a} \\
                \includegraphics[width=9cm]{hw4/pr3_v_FDR_b}
                \includegraphics[width=9cm]{hw4/pr3_v_pFDR_b} \\
                \includegraphics[width=9cm]{hw4/pr3_v_FDR_c}
                \includegraphics[width=9cm]{hw4/pr3_v_pFDR_c} \\
                I think my mistake made my test way overly conservative in a way I don't quite understand.
\end{document}
